Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-02-26 main.tex
Change probability variable from p to P.
Version 3
Bump to 4.dev
Fixed possessive to plural.

\title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
Low Speeds When Subjected to Handlebar Perturbations\\Version 3.dev}

@@

Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs:
%
\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bernoulli}(p_{ij}) \textrm{.}
\end{align}
%
We evaluate our hypothesis using a multivariate logistic regression model that
represents the probability of this binary outcome. The log-odds of the
probability is then a linear function of our independent variables with
\(\beta\) being the intercept and \(\alpha_k\) the linear coefficients to the
\(K\) independent variables \(x_{ij}^k\), i.e. all variables in
Table~\ref{tab:stat-model-variables} except \(f\).
%
\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) =
  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
  \label{eq:log-regress}
\end{align}

@@

To illustrate the effect of the balance assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the
coefficients in Table~\ref{tab:freq-coefs-6}. For simplicity's sake, the
interaction effects are not included. For example, assuming that the mean
angular impulse $\bar{L}$ of all the perturbations applied to a participant is
100~\si{\newton} and the standard deviation $\sigma_{L}=15$. The centered and
scaled angular impulse can be calculated by subtracting $\bar{L}$ from the
applied angular impulse $L$, and dividing this by $\sigma_{L}$. The same
applies for the perturbation order $j$, initial roll angle $\phi_0$, and
initial steer angle $\delta_0$. If we take the coefficients estimated for
cycling at 6~\si{\kph} and example means, the log-odds of falling can be
calculated as follows:
%
\begin{equation}
  \begin{split}
    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
    = & \quad
    \beta + \sum_{k=1}^{k=5}\alpha_k \frac{x_{ij}^{k}-\bar{x}_{ij}^{k}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 100\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64 s \\
    = &
    1.59 -0.64 s
  \textrm{.}
  \end{split}
\end{equation}
%
The state of the balance assist \(s\) is a binary variable. If the
balance assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{p_{ij}}{1-p_{ij}} & = e^{1.59-0.64s} =
  e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot e^{-0.64s} \\
  p_{ij}|_{s=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
  p_{ij}|_{s=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
\end{align}
%
Turning on the balance assist system reduces the probability that the
perturbation results in a fall from 83\% to 72\% in this case.

@@

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph} the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exiting the treadmill width or
placing a foot on the belt. On the other hand, cycle paths are a similar width
as the treadmill, so rider's are often limited in width when recovering from a
fall thus exiting the treadmill width may be an appropriate measure for
indicating a fall. While the treadmill simulates narrow cycle paths, real-world
paths may offer more lateral recovery space, providing riders with additional
opportunities to regain balance after a perturbation. However, testing in such
narrow conditions is still highly relevant, as system design and validation
should focus on extreme conditions like narrow paths, because even in wider
paths, obstacles such as parked cars or barriers can limit lateral space.

>>>>

\title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev}

@@

Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(P_{ij}\) that a fall occurs:
%
\begin{align}
  f_{ij} | P_{ij} \sim \textrm{Bernoulli}(P_{ij}) \textrm{.}
\end{align}
%
We evaluate our hypothesis using a multivariate logistic regression model that
represents the probability of this binary outcome. The log-odds of the
probability is then a linear function of our independent variables with
\(\beta\) being the intercept and \(\alpha_k\) the linear coefficients to the
\(K\) independent variables \(x_{ij}^k\), i.e. all variables in
Table~\ref{tab:stat-model-variables} except \(f\).
%
\begin{align}
  \log \left(\frac{P_{ij}}{1-P_{ij}} \right) =
  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
  \label{eq:log-regress}
\end{align}

@@

To illustrate the effect of the balance assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the
coefficients in Table~\ref{tab:freq-coefs-6}. For simplicity's sake, the
interaction effects are not included. For example, assuming that the mean
angular impulse $\bar{L}$ of all the perturbations applied to a participant is
100~\si{\newton} and the standard deviation $\sigma_{L}=15$. The centered and
scaled angular impulse can be calculated by subtracting $\bar{L}$ from the
applied angular impulse $L$, and dividing this by $\sigma_{L}$. The same
applies for the perturbation order $j$, initial roll angle $\phi_0$, and
initial steer angle $\delta_0$. If we take the coefficients estimated for
cycling at 6~\si{\kph} and example means, the log-odds of falling can be
calculated as follows:
%
\begin{equation}
  \begin{split}
    \log \left( \frac{P_{ij}}{1-P_{ij}} \right)
    = & \quad
    \beta + \sum_{k=1}^{k=5}\alpha_k \frac{x_{ij}^{k}-\bar{x}_{ij}^{k}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 100\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64 s \\
    = &
    1.59 -0.64 s
  \textrm{.}
  \end{split}
\end{equation}
%
The state of the balance assist \(s\) is a binary variable. If the
balance assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{P_{ij}}{1-P_{ij}} & = e^{1.59-0.64s} =
  e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot e^{-0.64s} \\
  P_{ij}|_{s=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
  P_{ij}|_{s=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
\end{align}
%
Turning on the balance assist system reduces the probability that the
perturbation results in a fall from 83\% to 72\% in this case.

@@

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph} the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exiting the treadmill width or
placing a foot on the belt. On the other hand, cycle paths are a similar width
as the treadmill, so riders are often limited in width when recovering from a
fall thus exiting the treadmill width may be an appropriate measure for
indicating a fall. While the treadmill simulates narrow cycle paths, real-world
paths may offer more lateral recovery space, providing riders with additional
opportunities to regain balance after a perturbation. However, testing in such
narrow conditions is still highly relevant, as system design and validation
should focus on extreme conditions like narrow paths, because even in wider
paths, obstacles such as parked cars or barriers can limit lateral space.
